---
title: "paper"
author: "Jacob K. Harvey"
date: "18/05/2021"
output:
  html_document: 
    css: "../documents/thesis_style.css"
bibliography: bibliography.json
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(results = FALSE)
options(digits = 2, scipen = -2)
```

## Comparison of specialization models
```{r}
library(readr)
odonates <- read_csv("datasets_derived/odonate_summaries_asv.csv")
source("scripts/specialization_gam_models.r")

pm <- summary(phylo_model)
im <- summary(immune_model)
mm <- summary(mass_model)
am <- summary(abun_model)
```
|Model                   |Estimate             |p-value           |
|:-----------------------|:--------------------|:-----------------|
|Host Phylogeny          |??                   |$`r pm$s.pv`$     |
|Log Host Immune Response|$`r im$p.coeff[[2]]`$|$`r im$p.pv[[2]]`$|
|Log Host Mass           |$`r mm$p.coeff[[2]]`$|$`r mm$p.pv[[2]]`$|
|Log Host Abundance      |$`r am$p.coeff[[2]]`$|$`r am$p.pv[[2]]`$|

## Sensitivity to phylogenetic v. non-phylogenetic indices (host range)
```{r}
pmnh <- summary(phylo_model_nhost)
imnh <- summary(immune_model_nhost)
mmnh <- summary(mass_model_nhost)
amnh <- summary(abun_model_nhost)
```
|Model                   |Estimate               |p-value             |
|:-----------------------|:----------------------|:-------------------|
|Host Phylogeny          |??                     |$`r pmnh$s.pv`$     |
|Log Host Immune Response|$`r imnh$p.coeff[[2]]`$|$`r imnh$p.pv[[2]]`$|
|Log Host Mass           |$`r mmnh$p.coeff[[2]]`$|$`r mmnh$p.pv[[2]]`$|
|Log Host Abundance      |$`r amnh$p.coeff[[2]]`$|$`r amnh$p.pv[[2]]`$|

# Sensitivity to sequence clustering
```{r}
odonates <- read_csv("datasets_derived/odonate_summaries_otu97.csv")
source("scripts/specialization_gam_models.r")

pm97 <- summary(phylo_model)
im97 <- summary(immune_model)
mm97 <- summary(mass_model)
am97 <- summary(abun_model)

odonates <- read_csv("datasets_derived/odonate_summaries_otu90.csv")
source("scripts/specialization_gam_models.r")

pm90 <- summary(phylo_model)
im90 <- summary(immune_model)
mm90 <- summary(mass_model)
am90 <- summary(abun_model)
```
|Model                   |Sequence similarity (%)|Estimate               |p-value             |
|:-----------------------|:----------------------|:----------------------|:-------------------|
|Host Phylogeny          |100                    |??                     |$`r pm$s.pv`$       |
|Log Host Immune Response|100                    |$`r im$p.coeff[[2]]`$  |$`r im$p.pv[[2]]`$  |
|Log Host Mass           |100                    |$`r mm$p.coeff[[2]]`$  |$`r mm$p.pv[[2]]`$  |
|Log Host Abundance      |100                    |$`r am$p.coeff[[2]]`$  |$`r am$p.pv[[2]]`$  |
|Host Phylogeny          |97                     |??                     |$`r pm97$s.pv`$     |
|Log Host Immune Response|97                     |$`r im97$p.coeff[[2]]`$|$`r im97$p.pv[[2]]`$|
|Log Host Mass           |97                     |$`r mm97$p.coeff[[2]]`$|$`r mm97$p.pv[[2]]`$|
|Log Host Abundance      |97                     |$`r am97$p.coeff[[2]]`$|$`r am97$p.pv[[2]]`$|
|Host Phylogeny          |90                     |??                     |$`r pm90$s.pv`$     |
|Log Host Immune Response|90                     |$`r im90$p.coeff[[2]]`$|$`r im90$p.pv[[2]]`$|
|Log Host Mass           |90                     |$`r mm90$p.coeff[[2]]`$|$`r mm90$p.pv[[2]]`$|
|Log Host Abundance      |90                     |$`r am90$p.coeff[[2]]`$|$`r am90$p.pv[[2]]`$|


![](../figures/pscale_phylo_plot_asv.svg)
<p class = "caption">
Distribution of mite ASVs across odonate phylogeny by mite specialization
</p>

![](../figures/pscale_phylo_plot_otu97.svg)
<p class = "caption">
Distribution of mite OTUs (97% similarity) across odonate phylogeny by mite specialization
</p>

![](../figures/pscale_phylo_plot_otu90.svg)
<p class = "caption">
Distribution of mite OTUs (90% similarity) across odonate phylogeny by mite specialization
</p>
